---
title: "CVD MR"
output:
  word_document: default
  html_document: default
date: "2024-02-19"
---

Full MR script for one exposure and multiple outcomes.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = NULL, warning = FALSE, message = FALSE) # nolint

# Try MR
source("Scripts/setup.R")

# get version of coloc
packageVersion("coloc")
packageVersion("TwoSampleMR")

# required columns for outcome: rsid, beta, se, eaf, effect_allele_col, other_allele_col, pval
```


The next two chunks are for reading in the exposure data and the outcome data. Add your own file names.

```{r eval = FALSE}
# # read in exposure data, PGC DEP
# exposure_data = read_exposure_data(
#   filename = here("Data", "DEP", "pgc-mdd2022-full-eur-v3.49.24.11.pgc"),
#   sep = "\t",
#   snp_col = "ID",
#   beta_col = "BETA",
#   se_col = "SE",
#   eaf_col = "FCON",
#   effect_allele_col = "A1",
#   other_allele_col = "A2",
#   log_pval = FALSE,
#   pval_col = "PVAL",
#   chr_col = "CHR",
#   pos_col = "BP"
# )
# exposure_data$exposure <- "DEP"
# ```
# 
# 
# ```{r eval = FALSE}
# cvd_traits <- c("ALLSTROKE", "IS", "CES", "LAS", "SVD", "CAD", "CAC", "CIMT")
#
# outcome_list <- list()
#
# for (i in cvd_traits) {
#
#   # read in outcome data, CVD
#   file <- paste0(i, "_with_RSID.tsv")
#
#   outcome_data = read_outcome_data(
#     filename = here("sumstats", i, file),
#     sep = "\t",
#     snps = exposure_data$ID, # shouldn't this be exposure_data$SNP?
#     snp_col = "RSID",
#     beta_col = "beta",
#     se_col = "standard_error",
#     eaf_col = "effect_allele_frequency",
#     effect_allele_col = "effect_allele",
#     other_allele_col = "other_allele",
#     pval_col = "p_value",
#     #ncase_col = "NCAS",
#     #ncontrol_col = "NCON",
#     log_pval = FALSE,
#     chr_col = "CHR",
#     pos_col = "BP"
#   )
#   outcome_data$outcome <- i
#
#   outcome_list[[i]] <- outcome_data
#
# }
#
# outcome_list
# save(outcome_list, file = here("sumstats", "outcome_forMR.RData"))
```

Preprocess the data: The exposure data is filtered and clumped to identify instruments. Then one harmonised dataframe is created from exposure and outcome.


```{r preprocessing, eval = FALSE}
# # exposure is always from exposure_data, but outcome_data comes from outcome_list
#
# SNP_numbers <- c() # store how many SNPs are left after each filtering step
# preprocessed_list <- list()
#
# for ( i in cvd_traits){
#   outcome_data <- outcome_list[[i]]
#
#   # Keep only SNPs that are in both datasets
#   # !! careful: refresh outcome data on every iteration but keep original exposure_data object the same
#   exposure_data_small <- exposure_data[exposure_data$SNP %in% outcome_data$SNP,]
#   outcome_data <- outcome_data[outcome_data$SNP %in% exposure_data_small$SNP,]
#
#   # Filtering and clumping in the exposure data
#   # pvalue filter
#   exposure_data_topsnps = exposure_data_small[exposure_data_small$pval.exposure < 5e-08, ]
#   # maf filter
#   exposure_data_topsnps_maf = exposure_data_topsnps %>% filter(eaf.exposure < 0.99 & eaf.exposure > 0.01)
#
#   # clumping
#   # using own function wrappers of plink clumping to adapt to using local ld files rather than server
#   # plink_bin <- "C:/Users/emma/AppData/Local/R/win-library/4.3/genetics.binaRies/bin/plink2.exe"
#   # bfile = here("ld_files", "g1000_eur")
#   exposure_data_clumped <- clump_data_locally( # defaults are clump_kb = 10000, clump_r2 = 0.001
#     dat = exposure_data_topsnps_maf,
#     bfile = "C:/Users/emma/Documents/git/ld-files/g1000_eur",
#     plink_bin = "C:/Users/emma/AppData/Local/R/win-library/4.3/genetics.binaRies/bin/plink2.exe"
#   )
#   # for server version use the TwoSampleMR version clump_data(). only required input argument is dat = your dataset
#
#   harmonised_data <- harmonise_data(exposure_dat = exposure_data_clumped, outcome_dat = outcome_data)
#
#   # Store some info about filtering process
#   SNP_numbers <- c()
#   SNP_numbers["beforefilter"] <- dim(exposure_data_small)[1]
#   SNP_numbers["pval"] <- dim(exposure_data_topsnps)[1]
#   SNP_numbers["maf"] <- dim(exposure_data_topsnps_maf)[1]
#   SNP_numbers["clumped"] <- dim(exposure_data_clumped)[1]
#   SNP_numbers["harmonised"] <- dim(harmonised_data)[1]
#
#   # And return something
#   preprocessed_list[[i]] <- list(data = harmonised_data, log = SNP_numbers)
# }
#
# preprocessed_list
# save(preprocessed_list, file = here("Int-results", "all_processed.RData"))
```

Run the main MR analysis and also all the essential follow-up analyses.

As described by Yuri:  

"First, a series of univariable 2SMR analyses were performed based on
the inverse variance weighted (IVW) estimator (Burgess et al., 2013),
pooling SNP-exposure/SNP-outcome estimates inversely weighted by
their standard error. Since IVW assumes that all SNPs are valid instruments or that the sum of the directional bias is zero, the robustness of
significant results was tested in sensitivity analyses based on weighted
median and MR-Egger estimators. The weighted median (Bowden et al.,
2016) estimator is the median of the weighted empirical distribution
function of individual SNP ratio estimates, providing consistent effect
estimates even if half of the instruments are invalid. 
The MR-Egger regression (Bowden et al., 2015) consists of a weighted linear regression similar to IVW relying on the InSIDE assumption (the magnitude of any pleiotropic effects should not correlate with the magnitude of the  main effect), 
providing valid effect estimate even if all SNPs are invalid
instruments under the ‘NOME’ assumption (uncertainty in the SNPexposure association estimates is negligible) (Bowden et al., 2017). 
At least 10 genetic instruments are recommended (Bowden et al., 2015) to
run adequately powered MR-Egger analyses. Furthermore, heterogeneity among included SNPs was tested via Cochran’s Q test, single SNP,
and leave-one-out SNP analyses. The presence of potential horizontal 
pleiotropy (a genetic instrument for exposure influencing the outcome
by mechanisms other than exposure) was tested using the MR-Egger
intercept (Bowden et al., 2017) and the MR-PRESSO (pleiotropy residual sum and outlier) method (Verbanck et al., 2018) (supplemental
methods). Finally, we performed reversed univariable 2SMR analyses
testing the potential causal impact of depression liability on AC circulating levels."

```{r results, eval = FALSE}
# #load(here("sumstats", "all_processed.RData"))
# main_results_mr <- list()
#
# for (i in names(preprocessed_list)) {
#   preprocessed_data <- preprocessed_list[[i]][["data"]]
#
#   # Run MR using various methods
#   mr_output <- mr(preprocessed_data, method_list = c("mr_ivw", 'mr_weighted_median', "mr_egger_regression"))
#
#   mr_or <- generate_odds_ratios(mr_output)
#   # to adjust for multiple testing using Benjamini-Hochberg procedure use FDR = p.adjust(pvalues, method="BH")
#   # https://www.r-bloggers.com/2023/07/the-benjamini-hochberg-procedure-fdr-and-p-value-adjusted-explained/
#   print(mr_or)
#
#   scatterplot <- mr_scatter_plot(mr_output, preprocessed_data)
#
#   # Instrument strength
#   fstatistic <- (preprocessed_data$beta.exposure^2) / (preprocessed_data$se.exposure^2) # approximation for sumstats
#   f_range <- c(min(fstatistic), max(fstatistic)) # F should not be < 10 (Pierce et al. 2011)
#
#   # Heterogeneity
#   mr_het <- mr_heterogeneity(preprocessed_data) # Chochran's Q
#   mr_leave1out <- mr_leaveoneout(preprocessed_data) # results for leave 1 out plot
#   oneout_plot <- mr_leaveoneout_plot(mr_leave1out) # leave 1 out plot
#   mr_1snp <- mr_singlesnp(preprocessed_data) # results for single SNP plot
#   onein_plot <- mr_forest_plot(mr_1snp) # single SNP plot
#
#   # Horizontal Pleiotropy
#   mr_pleiot <- mr_pleiotropy_test(preprocessed_data) # this gives the egger_intercept
#   print(mr_pleiot)
#   mr_presso_output <- run_mr_presso(preprocessed_data, NbDistribution = 1000, SignifThreshold = 0.05) # MR_presso - with outlier correction
#
#   main_results_mr[[i]] <- list(mr_output = mr_output, mr_or = mr_or, scatterplot = scatterplot
#                                , f_range = f_range, mr_het = mr_het, mr_leave1out = mr_leave1out
#                                , oneout_plot = oneout_plot ,mr_1snp = mr_1snp, onein_plot = onein_plot
#                                , mr_pleiot = mr_pleiot, mr_presso_output = mr_presso_output)
# }
#
# main_results_mr
# save(main_results_mr, file = here("Results", "MR_DEP-CVD_results.RData"))
```


# Main results

```{r}
load(here("Results", "MR", "MR_DEP-CVD_results.RData"))

# Nice table of main results
or_full_df <- as.data.frame(do.call("rbind", map(main_results_mr, "mr_or")))[c(3, 4, 5, 6, 12, 13, 14, 9)]
or_full_df$pval <- formatC(or_full_df$pval, format = "e", digits = 2) # pvalues to scientific notation
or_full_df <- or_full_df %>%
  rename("or.lci95" = "or_lci95", "or.uci95" = "or_uci95") %>% # change separator
  mutate(across(c(or, or.lci95, or.uci95), ~ round(.x, digits = 2))) # round to two digits

or_wide_df <- pivot_wider(
  data = or_full_df,
  names_from = method,
  names_vary = "slowest",
  names_glue = "{method}_{.value}",
  values_from = c(or, or.lci95, or.uci95, pval)
)

flextable(or_wide_df) %>%
  separate_header(split = "_") %>%
  fontsize(size = 14) %>%
  fit_to_width(8) %>%
  autofit()


# Special OR
mr_beta <- as.data.frame(do.call("rbind", map(main_results_mr, "mr_output"))) |>
  mutate(b = b * log(2))
special_or <- generate_odds_ratios(mr_beta)

head(special_or)
special_full_df <- special_or[c(3, 4, 5, 6, 12, 13, 14, 9)] |>
  rename("or.lci95" = "or_lci95", "or.uci95" = "or_uci95") |> # change separator
  mutate(across(c(or, or.lci95, or.uci95), ~ round(.x, digits = 2))) |> # round to two digits
  mutate(across(c(pval), ~ formatC(.x, format = "e", digits = 2)))

special_wide_df <- pivot_wider(
  data = special_full_df,
  names_from = method,
  names_vary = "slowest",
  names_glue = "{method}_{.value}",
  values_from = c(or, or.lci95, or.uci95, pval)
) |>
  mutate(`Inverse variance weighted_pval.FDR` = p.adjust(`Inverse variance weighted_pval`, method = "BH")) |> # adjust for multiple testing
  relocate(`Inverse variance weighted_pval.FDR`, .after = `Inverse variance weighted_pval`) |>
  mutate(`Weighted median_pval.FDR` = p.adjust(`Weighted median_pval`, method = "BH")) |> # adjust for multiple testing
  relocate(`Weighted median_pval.FDR`, .after = `Weighted median_pval`) |>
  mutate(`MR Egger_pval.FDR` = p.adjust(`MR Egger_pval`, method = "BH")) |> # adjust for multiple testing
  relocate(`MR Egger_pval.FDR`, .after = `MR Egger_pval`)

head(special_wide_df)
write.csv(special_wide_df, file = here("Results", "MR", "MR_DEP-CVD_full_main_with_FDR.csv"), row.names = TRUE)

flextable(special_wide_df) %>%
  separate_header(split = "_") %>%
  fontsize(size = 14) %>%
  fit_to_width(8) %>%
  autofit()
```


Note that we didn't expect an effect for CES (mostly genes specific to heart development) and indeed the estimate is not significant.


```{r}
# load(here("Results", "CVD_mr_results.RData"))
#
# # Nice table of main results
# full_df <- as.data.frame(do.call("rbind", map(main_results_mr, "mr_output")))[c(3,4,5,6,7,8,9)]
# or_full_df$pval <- formatC(or_full_df$pval, format = "e", digits = 2) # pvalues to scientific notation
# or_full_df <- or_full_df %>%
#   rename("or.lci95" = "or_lci95", "or.uci95" = "or_uci95") %>% # change separator
#   mutate(across(c(or, or.lci95, or.uci95), ~ round(.x, digits = 2))) # round to two digits
#
# or_wide_df <- pivot_wider(
#   data = or_full_df,
#   names_from = method,
#   names_vary = "slowest",
#   names_glue = "{method}_{.value}",
#   values_from = c(or, or.lci95, or.uci95, pval)
# )
#
# flextable(or_wide_df) %>% separate_header(split = "_")
```

```{r fig.dim = c(8, 17)}
# Plot collection
# plots <- unlist(map(main_results_mr, "scatterplot"))
p1 <- main_results_mr$ALLSTROKE$scatterplot$v3kGW8.7mNsA8
p2 <- main_results_mr$IS$scatterplot$v3kGW8.gYxmey
p6 <- main_results_mr$CAD$scatterplot$v3kGW8.AEKi7w
p4 <- main_results_mr$CES$scatterplot$v3kGW8.a4pcPh
p5 <- main_results_mr$LAS$scatterplot$v3kGW8.bpulR3
p3 <- main_results_mr$SVD$scatterplot$v3kGW8.KeIQy5
p7 <- main_results_mr$CAC$scatterplot$e7VlWv.JiGy5p
p8 <- main_results_mr$CIMT$scatterplot$e7VlWv.fv9RyK

p1 + p2 + p3 + p4 + p5 + p6 + p7 + p8 +
  plot_layout(ncol = 2, guides = "collect") &
  theme(legend.position = "bottom")
```

# MR more analyses

## Instrument strength

```{r }
# # Test instrument strength
# fstatistic <- (harmonised_data$beta.exposure^2) / (harmonised_data$se.exposure^2) # approximation for sumstats
# min(fstatistic) # F should not be < 10 (Pierce et al. 2011)
# max(fstatistic)

fstats <- as.data.frame(do.call("rbind", map(main_results_mr, "f_range")))
fstats <- cbind(or_wide_df$nsnp, round(fstats, 4))
colnames(fstats) <- c("N SNPs", "F-statistics_min", "F-statistics_max")

flextable(fstats %>% rownames_to_column(var = "Outcome")) %>% separate_header()
write.csv(fstats, file = here("Results", "MR", "MR_DEP-CVD_Fstats.csv"), row.names = TRUE)
fstats
```

Probably these make sense..? There is large overlap in the SNPs involved for each outcome, so the SNP with the weakest and the SNP with the strongest effect on the exposure could feasibly be the same for all stroke outcomes. 

# Heterogeneity

## Cochran's Q
A significant Cochran's Q indicates presence of an outlier.

```{r }
# # Heterogeneity
# mr_het <- mr_heterogeneity(harmonised_data) # this give Cochran's Q (if significant indicates outlier), one for Egger + one for IVW
qstats <- as.data.frame(do.call("rbind", map(main_results_mr, "mr_het")))
qstats["Q"] <- round(qstats["Q"])
qstats["df"] <- qstats["Q_df"]
qstats["pvalue"] <- formatC(qstats$Q_pval, format = "e", digits = 2)
qstats <- cbind(qstats[c("outcome", "exposure", "method", "Q", "df", "pvalue")])

q_wide <- pivot_wider(
  data = qstats,
  names_from = method,
  names_vary = "slowest",
  names_glue = "{method}_{.value}",
  values_from = c(Q, df, pvalue)
)

flextable(q_wide) %>%
  separate_header(split = "_") %>%
  add_header_row(values = c("", "Cochran's Q"), colwidths = c(2, 6))
```


## Leave one out analysis

```{r }
map(main_results_mr, "oneout_plot")
```

## Single SNP analysis

```{r }
map(main_results_mr, "onein_plot")
```

# Horizontal Pleiotropy

## MR-Egger Intercept

```{r }
# Pleiotropy
mr_pleiot <- as.data.frame(do.call("rbind", map(main_results_mr, "mr_pleiot")))
mr_pleiot[c("egger_intercept", "se")] <- round(mr_pleiot[c("egger_intercept", "se")], 5)
mr_pleiot["pval"] <- formatC(mr_pleiot$pval, format = "e", digits = 2)

flextable(mr_pleiot[-c(1, 2)])
```

## MR presso

```{r}
# MR_presso
mr_presso <- map(main_results_mr, c("mr_presso_output"))

outlier_corrected <- do.call(
  "rbind",
  lapply(mr_presso, function(x) {
    x[[1]][["Main MR results"]][2, c("Causal Estimate", "Sd", "P-value")]
  })
)

global_test <- as.data.frame(do.call(
  "rbind",
  lapply(mr_presso, function(x) {
    unlist(x[[1]][["MR-PRESSO results"]][["Global Test"]])
  })
))

distortion_coefficient <- as.data.frame(do.call(
  "rbind",
  lapply(mr_presso, function(x) {
    x[[1]][["MR-PRESSO results"]][["Distortion Test"]][["Distortion Coefficient"]]
  })
))
distortion_pvalue <- as.data.frame(do.call(
  "rbind",
  lapply(mr_presso, function(x) {
    unlist(x[[1]][["MR-PRESSO results"]][["Distortion Test"]][["Pvalue"]])
  })
))


mr_presso_table <- cbind(or_wide_df$exposure, or_wide_df$nsnp, global_test, outlier_corrected, distortion_coefficient, distortion_pvalue)
colnames(mr_presso_table) <- c("Exposure", "N SNPs", "GlobalTest_RSS", "GlobalTest_pvalue", "OutlierCorrected_estimate", "OutlierCorrected_se", "OutlierCorrected_pvalue", "DistortionTest_Coefficient", "DistortionTest_pvalue")
mr_presso_table$OutlierCorrected_pvalue <- formatC(mr_presso_table$OutlierCorrected_pvalue, format = "e", digits = 2)
mr_presso_table$GlobalTest_RSS <- round(as.numeric(mr_presso_table$GlobalTest_RSS), 3)


flextable(mr_presso_table %>% rownames_to_column(var = "Outcome")) %>%
  separate_header() %>%
  fit_to_width(8)
```



